Detailed Chemical Abundances with \tc{}

### 2015-10-06 article/ms.tex
Mostly word-wrapping, with some minor text changes

\begin{abstract}
More than 15 million stellar spectra are expected to be acquired by ground-based surveys in
the coming decade. The detailed chemical abundance analysis of these data poses
a significant challenge. \tc{} provides a purely data-driven
approach to stellar parameter (label) determination. Cross-validation tests
demonstrate that the precision achievable is comparable to existing methods that
rely on model atmospheres and spectral synthesis, but \tc{} is $\sim$300,000 times
faster, making it perfectly applicable for surveys. Including individual abundances by extending \tc{} dimensionality can create enormous model difficulties, severely limiting the interpretability of the results. Here
we describe an alternative approach to including individual element abundances
in \tc{}. Using modest simplifications of the physics of spectral line formation,
we demonstrate precisely \textit{why} an approach like \tc{} works so effectively,
and provide a physically-motivated model that describes flux values, given the
stellar parameters. The inclusion of individual chemical abundances does not
impact on model flexibility, allowing for simultaneous training on detailed
chemical abundances (which are formed through well-understood interactions), as well as arbitrary labels where the impact on the spectra
is unknown or poorly understood (e.g., masses, ages). Crucially, we provide a direct link
between data-driven models and theory of model atmospheres and spectral line formation. We demonstrate that
detailed chemical abundances can be measured with a precision of 0.0X-0.0Y~dex
in Z APOGEE spectra on a modern CPU in just one minute. We present abundance
measurements of 15 elements in Z APOGEE spectra from DR12.
\end{abstract}

@@

Ground-based stellar spectroscopic surveys have enormous scientific potential. Key absorption features in stellar spectra permit inferences on the internal structure of stars and their detailed chemical composition. Given a large enough sample of stars, these inferences can directly guide or constrain models over a large range of physical scales: the formation of planets, stars and clusters, supernovae yields and nucleosynthetic pathways, as well as the large scale evolution of the Milky Way. For these reasons an increasing number of stellar spectroscopic surveys are being undertaken, with each survey generation increasing in sample size \citep{gaia_eso; galah; lamost; apogee; 4most}.

% Data have outgrown the methods.
In some ways, the spectroscopic data have outgrown the methods used to analyse them. High-resolution ($\mathcal{R} > 20,000$) stellar spectroscopy is colloquially considered to be a `dark art': analyses are carefully performed by an expert spectroscopist, who themselves were trained by an expert in exactly where the stellar continuum should be placed, and which atomic lines are considered 'clean' or 'dirty'. This work is equally impressive as it is subjective: the best spectroscopists host an encyclopaedic knowledge of atomic absorption lines with decades of experience, yet in blind-tests their decisions of where to place the continuum quickly becomes the most dominant source of systematic error.
% Lot of unextracted information in these spectra (e.g., mass, ages)

% Chemical abundance analysis takes time.
Classical spectroscopic analyses are time-intensive both in human and computing resources. The Gaia-ESO Survey has been particularly instrumental in minimising the human resources required for analysis: all analysis codes of FGK stars employed in Gaia-ESO are now completely automatised. The computing challenge remains: some analysis codes take up to one hour (single core) to determine stellar parameters (\teff, \logg, \mh) for one star. Given the data scale (e.g., $\sim$400 stars observed simultaneously with HERMES), this poses a modest computational problem.

% These analysis techniques are not optimal, either.
These analyses also have strict limitations in validity, and a range of options are available. The community standard is to perform chemical abundance analyses using 1D model photospheres under the assumption of local thermal equilibrium (LTE). There are a number of photospheric models available. Detailed information about the species (atomic and molecular) opacities is required, and partition functions are usually approximated from polynomials. Atomic data (e.g., oscillator strengths) must be well-known, and the treatment of Stark and XX broadening must be carefully included. Results are further sensitive to the manner in which the equation of state was solved. Simply put, there are many things to consider for a careful analysis.

@@

This form is flexible: the label vector $\lv$ may be linear in just a few common
labels (e.g., \teff, \logg, \mh\footnote{\citet{Ness15a} use [Fe/H] in $\lv$, but for the 
sake of distinguishing between overall metallicity and individual abundances,
throughout this \article{} we opt for \mh to describe overall metallicity})
or a complex combination of all $K$ available labels, some of which may have an
 unknown or non-trivial relationship with spectral fluxes \citep[e.g., ages, 
 masses, see][]{Nissen15, Ness15b}. The underlying principle
is that the flux for any pixel from any star is predictable with a simple (usually
 a polynomial) model, given the labels. Indeed, \citet{Ness15a}
employ a quadratic-in-labels spectral model using stellar
parameters ($T_{\rm eff}$, $\log{g}$, [M/H])
and demonstrate through cross-validation that these labels can be 
measured with a precision of 95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, 0.08 in 
[M/H], with negligible biases.

The noise (Eq. \ref{eq:cannon}) can be taken as a quadrature combination of the pixel uncertainty variance $\sigma_{n\lambda}^2$ and the intrinsic variance of the spectral model at each wavelength $s_\lambda^2$. The flux $f_{n\lambda}$ can be written as a linear function of the label vector, leading to a single-pixel log-likelihood function

@@

Given some sample of stars with known labels (the training set), the label vector $\lv$ and fluxes $f_{n\lambda}$ are known, allowing us to solve for the parameters $[\cv,s_\lambda^2]$ for each pixel with wavelength $\lambda$. Here we will optimise the negative log-likelihood (Equation \ref{eq:pixel-log-likelihood}). This is the training step of \tc{}, where the parameters $[\cv,s_\lambda^2]$ are optimised for a single pixel at $\lambda$ by considering all stars $N$:

\begin{equation}
\cv,s_\lambda \leftarrow^{\arg\,\max} {\cv, s_\lambda} \sum_{n=1}^N \ln{p}\left(f_{n\lambda}\given\cv^\intercal,\lv,s_\lambda^2\right)
\end{equation}

Once the coefficients $[\cv,\ssq]$ are known, \tc{} provides a \textit{generative model} for spectral fluxes. Through a cheap linear product of $\cv^\intercal$ and $\lv$, \tc{} is approximating spectral synthesis, an otherwise extremely expensive endeavour. We refer the reader to \citet{Ness15a} for a more detailed description of \tc{}.

Before extending \tc{} to include individual chemical abundances, it is
a useful exercise to understand \textit{why} a simple spectral model can 
approximate synthesis calculations with comparable fidelity, and thus \textit{how} this framework can be extended to include individual abundances. At present the applications of \tc{} have been restricted to stellar spectra with resolving powers of $\mathcal{R} = 22,500$ \citep[APOGEE;][]{Ness15a} or less \citep[e.g., LAMOST;][]{Ho15}. In these regimes, the shapes of unsaturated spectral lines are well-represented by Gaussian profiles. This fact leads to several critical insights, which we describe below.

In stellar spectroscopy we are frequently presented with the idea of the equivalent width (EW; $\mathcal{W}$) of an absorption line. Consider a single (unblended) atomic line at rest wavelength $\lambda_r$, where the flux values at and surrounding $\lambda_r$ ($f_{n\lambda_r-{\delta}},f_{n\lambda_{r+\delta}}$) are well-approximated by a \textit{Cannon} spectral model. Because the absorption line is well-represented by a Gaussian profile at this spectral resolution, the flux at $\lambda_r$ is related to the EW of the line through

@@

where $\sigma$ is the standard deviation of the profile, which is well-approximated by width of the line spread function (LSF) kernel. That is to say, for resolutions $\mathcal{R} \lesssim 20,000$, $\sigma$ is dominated by the spectrograph, not the absorption line or the labels (stellar parameters). If we presume that the LSF is well-known, or reasonably approximated by a fixed spectral resolution $\mathcal{R}$, then it is clear to see that \tc{} is implicitly modelling the EW of an atomic line as a function of stellar labels, and modelling spectral flux through a per-pixel scaling of $\sigma$. Indeed, the success of \tc{} suggests that the EW of an atomic line can also be well-approximated by a simple model relating the labels (i.e., stellar parameters and abundance of a single species). If such a model can be constructed and appropriately included within the existing \textit{Cannon} framework, it would permit the determination of detailed chemical abundances, whilst maintaining the flexibility in training on labels with poorly-known relationships on spectral fluxes (e.g., ages).

We note that Equation \ref{eq:ew-flux} does not imply that \tc{} is limited in applicability to low-resolution ($\mathcal{R} \lesssim 20,000$) spectra. Quite the opposite. \tc{} is sufficiently flexible that in extremely high-resolution spectra, where absorption lines are more closely represented by Lorentzian or Hertzting functions, the flux values $f_{n\lambda}$ in the wings of a strong line are presumably still well-modelled by \tc{}. Indeed, the line $\mathcal{W}$ would still be predictable with high-fidelity, implying that the coefficients $\cv$ for pixels in the wings just scale in a different way. Equation \ref{eq:ew-flux} allows us to directly translate $\mathcal{W}$ to $f_{n\lambda}$, thereby accounting for the flux contribution from a single atomic line. If the spectral resolving power were sufficiently high such that the absorption profiles appeared Lorentzian in nature, some prior information on the shape $\gamma$ would be required to translate $\mathcal{W}$ to $f_{n\lambda}$. Thus, while Equation \ref{eq:ew-flux} allows the translation from $\mathcal{W}$ to $f_{n\lambda}$ for $\mathcal{R} \lesssim 20,000$ spectra, given some knowledge of the LSF, no limiting interpretation about \tc{} should be drawn. Importantly, all future stellar spectroscopic surveys  currently expect $\mathcal{R} \lesssim 20,000$, in the perfect Gaussian-profile regime for this extension of \tc{}.

>>>>

\begin{abstract}
More than 15 million stellar spectra will be acquired by ground-based surveys in
the coming decade. The detailed chemical abundance analysis of these data poses
a significant challenge. \tc{} provides a purely data-driven
approach to stellar parameter (label) determination. Cross-validation tests
demonstrate that the precision achievable is comparable to existing methods that
rely on model atmospheres and spectral synthesis, but \tc{} is $\sim$300,000 times
faster, making it perfectly applicable for surveys. Including individual abundances by extending \tc{} dimensionality can create enormous model difficulties, severely limiting the interpretability of the results. Here
we describe an alternative approach to including individual element abundances
in \tc{}. Using modest simplifications of the physics of spectral line formation,
we demonstrate precisely \textit{why} an approach like \tc{} works so effectively,
and provide a physically-motivated model that describes flux values, given the
stellar parameters. The inclusion of individual chemical abundances does not
impact on model flexibility, allowing for simultaneous training on detailed
chemical abundances (which are formed through well-understood interactions), as well as arbitrary labels where the impact on the spectra
is unknown or poorly understood (e.g., masses, ages). Crucially, we provide a direct link
between data-driven models and theory of model atmospheres and spectral line formation. We demonstrate that
detailed chemical abundances can be measured with a precision of 0.0X-0.0Y~dex
in Z APOGEE spectra on a modern CPU in just one minute. We present abundance
measurements of 15 elements in Z APOGEE spectra from DR12.
\end{abstract}

@@

Ground-based stellar spectroscopic surveys have enormous scientific potential.
  Key absorption features in stellar spectra permit inferences on the internal 
structure of stars and their detailed chemical composition.  Given a sufficiently
large sample of stars, these inferences can directly guide or constrain models 
over a large range of physical scales: the formation of planets, stars and 
clusters, supernovae yields and nucleosynthetic pathways, as well as the large 
scale evolution of the Milky Way.  For these reasons an increasing number of 
stellar spectroscopic surveys are being undertaken, with each survey generation 
increasing in sample size \citep{gaia_eso; galah; lamost; apogee; 4most}.

% Data have outgrown the methods.
In some ways, the spectroscopic data have outgrown the methods used to analyse
them. High-resolution ($\mathcal{R} > 20,000$) stellar spectroscopy has 
colloquially been considered as a `dark art': analyses are carefully performed 
by an expert spectroscopist, who themselves were trained by an expert in exactly
where the stellar continuum should be placed, and which atomic lines are 
considered `clean' or `dirty'. This work is equally impressive as it is 
subjective: the best spectroscopists host an encyclopaedic knowledge of atomic
absorption lines with decades of experience, yet in blind-tests their decisions 
of where to place the continuum quickly becomes the most dominant source of 
systematic error.
% Lot of unextracted information in these spectra (e.g., mass, ages)

% Chemical abundance analysis takes time.
Classical spectroscopic analyses are time-intensive both in human and computing
resources. The Gaia-ESO Survey has been particularly instrumental in minimising
the human resources required for analysis: all analysis codes of FGK stars 
employed in Gaia-ESO are now completely automatised.  However the computing
challenge remains: some analysis codes take up to one hour (single core) to 
determine stellar parameters (\teff, \logg, \mh) for a single star. Given the 
data scale (e.g., $\sim$400 stars observed simultaneously with HERMES), this 
poses a modest computational problem.

% These analysis techniques are not optimal, either.
These analyses also have strict limitations in validity, and a range of options
are available.  Many of these options are not straightforward.  The community 
standard is to perform chemical abundance analyses using 1D model photospheres 
under the assumption of local thermal equilibrium (LTE). There are a number of 
photospheric models available.  Detailed information about the species (atomic 
and molecular) opacities is required, and partition functions are usually 
approximated from polynomials. Atomic data (e.g., oscillator strengths) must be
well-known, and the treatment of Stark and XX broadening must be carefully
included. Results are further sensitive to the manner in which the equation of 
state was solved. Simply put, there are many things to consider for a careful 
analysis.

@@

This form is flexible: the label vector $\lv$ may be linear in just a few common
labels (e.g., \teff, \logg, \mh\footnote{\citet{Ness15a} use [Fe/H] in $\lv$, 
but for the sake of distinguishing between overall metallicity and individual 
abundances, throughout this \article{} we opt for \mh to describe overall 
metallicity}) or a complex combination of all $K$ available labels, some of which
may have an unknown or non-trivial relationship with spectral fluxes \citep[e.g.,
ages, masses, see][]{Nissen15, Ness15b}. The underlying principle is that the 
flux for any pixel from any star is predictable with a simple model, given the
labels. Indeed, \citet{Ness15a} employ a quadratic-in-labels spectral model 
using stellar parameters ($T_{\rm eff}$, $\log{g}$, [M/H]) and demonstrate 
through cross-validation that these labels can be measured with a precision of 
95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, 0.08 in [M/H], with negligible biases.

% TODO: This needs rewording:
The noise in Equation \ref{eq:cannon} can be taken as a quadrature combination
of the variance in pixel flux $\sigma_{n\lambda}^2$ and the intrinsic variance 
of the spectral model at each wavelength $s_\lambda^2$. The flux $f_{n\lambda}$
can be written as a linear function of the label vector, leading to the
single-pixel log-likelihood function

@@

The parameters $[\cv,s_\lambda^2]$ for each pixel with wavelength $\lambda$ need
to be inferred. This is accomplished given some sample of stars where the labels
are known with some fidelity (i.e., the training set). Here we will optimise the
negative log-likelihood (Equation \ref{eq:pixel-log-likelihood}). This is the 
training step of \tc{}, where the parameters $[\cv,s_\lambda^2]$ are optimised 
for a single pixel at $\lambda$ by considering all stars $N$:

% TODO: Need to re-place 'arg max'
\begin{equation}
\cv,s_\lambda \leftarrow^{\arg\,\max} {\cv, s_\lambda} \sum_{n=1}^N \ln{p}\left(f_{n\lambda}\given\cv^\intercal,\lv,s_\lambda^2\right)
\end{equation}

Once the coefficients $\{\cv,\ssq\}$ are known, \tc{} provides an approximate
\textit{generative model} for spectral fluxes. Through a cheap linear product of
$\cv^\intercal$ and $\lv$, \tc{} is approximating spectral synthesis, an 
otherwise extremely expensive endeavour. This provides a number of advantages,
the most straightforward of which is the ability to efficiently determine $\lv$
(i.e., the stellar parameters) for a new star outside the training set in a
fraction of a second, which is about five orders of magnitude faster than
existing analysis codes. We refer the reader to \citet{Ness15a} for a more
detailed description of \tc{}.

It is not straightforward to add individual chemical abundances as labels and
expect equally impressive results.  Some chemical abundances only affect few
pixels of the spectrum. Other elements only present transitions that are
frequently blended with other elements, implying covariance between elemental
abundances.  And all transitions will have linear (or higher-order) 
relationships with stellar parameters.  Addressing these issues by increasing
the complexity of the label vector introduces additional issues (e.g.,
over-fitting, interpretability).  Although the existing \textit{Cannon}
framework can be extended to include individual chemical abundances, they cannot
be introduced by extensions of the label vector without additional issues.

% TODO: rewrite this so it all flows better.
Before we extend \tc{} to include individual chemical abundances, it is a useful
 exercise to understand \textit{why} a simple spectral model can 
approximate synthesis calculations with comparable fidelity, and thus 
\textit{how} this framework can be extended to include individual abundances. At
present the applications of \tc{} have been restricted to stellar spectra with 
resolving powers of $\mathcal{R} = 22,500$ \citep[APOGEE;][]{Ness15a} or less 
\citep[e.g., LAMOST;][]{Ho15}. In these regimes, the shapes of unsaturated 
spectral lines are well-represented by Gaussian profiles. This leads to several 
crucial insights, which we describe below.

In stellar spectroscopy we are frequently presented with the idea of the 
equivalent width (EW; $\mathcal{W}$) of a spectral line. Consider a single 
(unblended) atomic line at rest wavelength $\lambda_r$, where the flux values 
surrounding $\lambda_r$ ($f_{n\lambda_r-{\delta}},f_{n\lambda_{r+\delta}}$) are 
well-approximated by a \textit{Cannon} spectral model. Because the absorption 
line is well-represented by a Gaussian profile at this spectral resolution, the 
flux at $\lambda_r$ is related to the EW of the line through

@@

where $\sigma$ is the standard deviation of the profile. For typical FGK-type
stars observed with $\mathcal{R} \lesssim 20,000$, $\sigma$ will be dominated by
the line spread function (LSF). If we presume that the LSF is well-known, or 
reasonably approximated by a fixed spectral resolution $\mathcal{R}$, then it is
clear to see that \tc{} is implicitly modelling the EW of an atomic line as a 
function of stellar labels, and modelling spectral flux through a per-pixel 
scaling of $\sigma$. Indeed, the success of \tc{} suggests that the EW of an 
atomic line can also be well-approximated by a simple model relating the labels 
(i.e., stellar parameters and abundance of a single species). If such a model 
could be constructed and appropriately included within the existing 
\textit{Cannon} framework, it would permit the determination of individual 
chemical abundances, whilst maintaining the flexibility in training on labels 
with poorly-known relationships on spectral fluxes (e.g., ages).

We note that Equation \ref{eq:ew-flux} does not imply that \tc{} is limited in
applicability to low-resolution ($\mathcal{R} \lesssim 20,000$) spectra. Quite 
the opposite. \tc{} is sufficiently flexible that in extremely high-resolution 
spectra, where absorption lines are more closely represented by Lorentzian or
Hertzting functions, the flux values $f_{n\lambda}$ in the wings of a strong 
line are still well-modelled by \tc{} (as judged by $\ssq$ and performance in
cross-validation tests). Indeed, the line $\mathcal{W}$ would still be 
predictable with high-fidelity, implying that the coefficients $\cv$ for pixels
in the wings just scale in a different way. Equation \ref{eq:ew-flux} allows us
to directly translate $\mathcal{W}$ to $f_{n\lambda}$, thereby accounting for 
the flux contribution from a single atomic line. If the spectral resolving power
were sufficiently high such that the absorption profiles appeared Lorentzian in
nature, some prior information on the shape $\gamma$ would be required to 
translate $\mathcal{W}$ to $f_{n\lambda}$. Thus, while Equation \ref{eq:ew-flux}
allows the translation from $\mathcal{W}$ to $f_{n\lambda}$ for 
$\mathcal{R} \lesssim 20,000$ spectra, given some knowledge of the LSF, no 
limiting interpretation about \tc{} should be drawn. Importantly, all future 
stellar spectroscopic surveys are currently planning for spectral resolutions
$\mathcal{R} \lesssim 20,000$, in the perfect Gaussian-profile regime for this
extension of \tc{}.
